\bf On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2024-11-24 Manuscript/manuscript.tex
minor

This paper addresses two primary objectives from a statistical
perspective, using modern methodologies to analyze historical
data. First, we investigate whether reaction times (RTs) at the 2022
World Championships were significantly different from other
competitions, focusing on athletes who competed in multiple
events. Using a matched-pairs design, we compare RTs across the 2022,
2019, and 2023 World Championships, as well as the 2022 national-level
competitions. This approach isolates the effect of the competition
year while controlling for individual performance. The clustered
nature of the data, with the goal of assessing differences across
events within athletes, was analyzed using a rank-based comparison
approach for clustered data \citep{datta2005rank}. Second, we evaluate
the appropriateness of the 0.1-second disqualification threshold by
modeling RTs from World Championships held from 1999 onward. A
generalized gamma distribution with random effects for both venue and
heat was applied within the framework of the generalized additive
model for location, scale, and shape (GAMLSS)
\citep{stasinopoulos2024generalized}. This model estimates the
probability of RTs falling below the threshold, providing a
statistical assessment of RT consistency and the validity of the
current threshold.

The rest of this paper is organized as follows. Section~\ref{sec:Data}
describes the data collection and datasets used for the two
objectives. Section~\ref{sec:rank} outlines the rank-based comparison
methods for objective~1, rank-based comparison for clustered data with
subunit-level gruping to analyze differences in RTs across different
competitions. Section~\ref{sec:gamlss} details the methods for
objective~2, the gernalized Gamma GAMLSS to evaluate the appropriatenes
of the 0.1-second threshold. Section~\ref{sec:Results} presents the
results of both objectives and discusses the implications.
Finally, Section~\ref{sec:concludingremarks} highlights the
paper’s impact, limitations, and potential for future research.
All data and code for our analysis are provided in the supplementary
materials.

@@

Based on an exploratory analysis, the RTs are adequately
modeled by a Generalized Gamma (GG) distribution with random effects in
model parameters. The GG distribution has three parameters, denoted by
$\text{GG}(\mu, \sigma, \nu)$ has density function
\begin{equation}
  \label{eq:gg}
f_Y(y \mid \mu, \sigma, \nu) =
\frac{|\nu| \theta^\theta z^{\theta}}{\Gamma(\theta) y}
\exp\left(-z \theta\right),
\end{equation}
for $y > 0$, $\mu > 0$, $\sigma > 0$, and $\nu \neq 0$,
where $z = \left(\frac{y}{\mu}\right)^\nu$,
$\theta = 1 / (\sigma^2 \nu^2)$, and
$\Gamma(\cdot)$ denotes the Gamma function.
The GG distribution is highly flexible, encompassing several
well-known distributions as special cases, such as the
Weibull ($\mu = \nu)$ and  Gamma $(\nu = 1)$ distributions.
Its mean is
\[
  \mu \frac{\Gamma(\theta + 1 / \nu)}
  {\theta^{1 / \nu} \Gamma(\theta)},
\]
provided $\theta > -1 / \nu$. Here,
$\mu$ scales the central tendency, $\sigma$ controls
dispersion, and $\nu$ determines skewness. This parameterization allows
the distribution to model asymmetric and heavy-tailed data effectively, making
it particularly suitable for RTs.
An implementation of this distribution is available from R package
\texttt{gamlss.dist} \citep{rigby2019distributions}.

@@

\begin{table}
  \centering
  \caption{Estimated fixed-effect parameters with standard errors in
    parentheses and estimated variance of the random effects from the
    fitted generalized Gamma distribution with venue level random
    effects in $\mu$ and heat level random effects in $\sigma$ in
    Models~\eqref{eq:gg}--\eqref{eq:sigma}.}
  \label{tab:ggfit}
  \begin{tabular}{c c c c c c}
    \toprule
    Data set & $\beta_0$ & $\gamma_0$ & $\nu$ & $\sigma_v$ & $\sigma_h$ \\
    \midrule
    Excluding 2022 & $-$1.910 (0.005)  & $-$2.200 (0.025) & $-$1.177 (0.442) & 0.043 & 0.326 \\
    Including 2022 & $-$1.910 & $-$2.200 & $-$1.178 & 0.058 & 0.320 \\
    \bottomrule
  \end{tabular}
\end{table}

The fitted parameters of the generalized Gamma distribution int he
GAMLSS framework in Equations~\eqref{eq:gg}--\eqref{eq:sigma} are
summarized in Table~\ref{tab:ggfit}. Results obtained from both
excluding and including 2022 data are reported. The fixed-effect
parameters include $\beta_0$, $\gamma_0$, and $\nu$, corresponding to
the intercept of the log-location, log-scale, and shape of the GG
distribution, respectively. Random effects account for variability at
the venue level $\sigma_v$ on the log-scale of the $\mu$ parameter
and at the heat level $\sigma_h$ on the log-scale of the $\sigma$
parameter in the density in Equation~\eqref{eq:gg}. The variance
of the venue-level random effect is smaller than the heat-level random
effect variance, suggesting that heat-level variability in the scale
parameter is substantial, though on the dispersion parameter. When the
2022 data is included, the fixed-effect estimates remain stable,
indicating robustness, but the venue-level random-effect variance
increases to 0.0576, reflecting additional variability in the location
parameter, while the heat-level random-effect variance remains stable.
decreases slightly to 0.3198. These results
highlight that RTs are influenced by both venue and heat-level
factors, with heat-level variability playing a more prominent role in the scale
parameter, and that the inclusion of 2022 introduces greater venue-level
variability, likely due to systematic differences in RTs that year.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{diagnosis.pdf}
  \caption{Diagnosis of the fitted generalized Gamma distribution with
    random effects in model parameters: kernerl density of 1 million
    observations drawn from the fitted model overlaid with the
    histogram of the observed RTs (left); Q-Q plot of the
    normal z-score of the quantile residuals from the fitted model.}
  \label{fig:diagnosis}
\end{figure}

Figure~\ref{fig:diagnosis} presents diagnostic checks for the fitted
generalized Gamma distribution model with random effects. The left
panel compares the kernel density estimate of one million simulated reaction
times from the fitted model to the histogram of the observed RTs.
The close alignment between the density curve and the histogram suggests that
the fitted GG model adequately captures the overall distribution of the
RTs. The right panel shows a Q-Q plot of the z-scores of the
quantile residuals from the fitted model. The points lie approximately along
the 45-degree reference line, indicating that the residuals are consistent
with the standard normal distribution, supporting the adequacy of the model
fit. These diagnostics collectively demonstrate that the fitted model provides
a reasonable representation of the observed RT data.

@@

The fitted generalized Gamma GAMLSS model with both venue- and heat-level
random effects provides a framework for assessing how extreme RTs
below certain thresholds are. The probability of observing a RT
below a given threshold, assuming no intentional false starts, was approximated
by generating 10 million realizations from the fitted model.
Table~\ref{tab:Sim_probability} summarizes the probabilities of observing
RTs below 0.08, 0.09, and 0.10 seconds under two scenarios: one
excluding and the other including data from 2022. Excluding 2022 slightly
reduces the probability of observing a fast RT, but the difference
is small. For example, the probability of a RT below 0.10 seconds
decreases from $2.76 \cdot 10^{-3}$ (approximately one in 362 starts) to
$1.94 \cdot 10^{-3}$ (approximately one in 515 starts) when 2022 is excluded.
Lowering the RT threshold from 0.10 to 0.08 seconds drastically
reduces the likelihood of observing a RT below the barrier, with
the probability dropping from one in every 362 starts (at 0.10 seconds) to one
in every 14620 starts (at 0.09 seconds) and one in every 146198 starts (at 0.08
seconds) when 2022 is included. These results highlight the rarity of extremely
fast RTs and substantiate the recommendations of \citet{komi2009iaaf}
to carefully consider the selection of RT thresholds.

>>>>

This paper addresses two primary objectives from a statistical
perspective, using modern methodologies to analyze historical
data. First, we investigate whether reaction times (RTs) at the 2022
World Championships were significantly different from other
competitions, focusing on athletes who competed in multiple
events. Using a matched-pairs design, we compare RTs across the 2022,
2019, and 2023 World Championships, as well as the 2022 national-level
competitions. This approach isolates the effect of the competition
year while controlling for individual performance. The clustered
nature of the data, with the goal of assessing differences across
events within athletes, was analyzed using a rank-based comparison
approach for clustered data \citep{datta2005rank}. Second, we evaluate
the appropriateness of the 0.1-second disqualification threshold by
modeling RTs from World Championships held from 1999 onward. A
generalized Gamma (GG) distribution with random effects for both venue and
heat was applied within the framework of the generalized additive
model for location, scale, and shape (GAMLSS)
\citep{stasinopoulos2024generalized}. This model estimates the
probability of RTs falling below the threshold, providing a
statistical assessment of RT consistency and the validity of the
current threshold.

The rest of this paper is organized as follows. Section~\ref{sec:Data}
describes the data collection and datasets used for the two
objectives. Section~\ref{sec:rank} outlines the rank-based comparison
methods for objective~1, rank-based comparison for clustered data with
subunit-level gruping to analyze differences in RTs across different
competitions. Section~\ref{sec:gamlss} details the methods for
objective~2, the GG model with random effects to evaluate the appropriatenes
of the 0.1-second threshold. Section~\ref{sec:Results} presents the
results of both objectives and discusses the implications.
Finally, Section~\ref{sec:concludingremarks} highlights the
paper’s impact, limitations, and potential for future research.
All data and code for our analysis are provided in the supplementary
materials.

@@

Based on an exploratory analysis, the RTs are adequately
modeled by a GG distribution with random effects in
model parameters. The GG distribution has three parameters, denoted by
$\text{GG}(\mu, \sigma, \nu)$ has density function
\begin{equation}
  \label{eq:gg}
f_Y(y \mid \mu, \sigma, \nu) =
\frac{|\nu| \theta^\theta z^{\theta}}{\Gamma(\theta) y}
\exp\left(-z \theta\right),
\end{equation}
for $y > 0$, $\mu > 0$, $\sigma > 0$, and $\nu \neq 0$,
where $z = (y / \mu)^\nu$,
$\theta = 1 / (\sigma^2 \nu^2)$, and
$\Gamma(\cdot)$ denotes the Gamma function.
The GG distribution is highly flexible, encompassing several
well-known distributions as special cases, such as the
Weibull ($\mu = \nu)$ and  Gamma $(\nu = 1)$ distributions.
Its expectation is
\[
  \frac{\mu \Gamma(\theta + 1 / \nu)}
  {\theta^{1 / \nu} \Gamma(\theta)},
\]
provided $\theta > -1 / \nu$. Here,
$\mu$ scales the central tendency, $\sigma$ controls
dispersion, and $\nu$ determines skewness. This parameterization allows
the distribution to model asymmetric and heavy-tailed data effectively, making
it particularly suitable for RTs.
An implementation of this distribution is available from R package
\texttt{gamlss.dist} \citep{rigby2019distributions}.

@@

\begin{table}
  \centering
  \caption{Estimated fixed-effect parameters with standard errors in
    parentheses and estimated variance of the random effects from the
    fitted GG distribution with venue level random
    effects in $\mu$ and heat level random effects in $\sigma$ in
    Models~\eqref{eq:gg}--\eqref{eq:sigma}.}
  \label{tab:ggfit}
  \begin{tabular}{c c c c c c}
    \toprule
    Data set & $\beta_0$ & $\gamma_0$ & $\nu$ & $\sigma_v$ & $\sigma_h$ \\
    \midrule
    Excluding 2022 & $-$1.910 (0.005)  & $-$2.200 (0.025) & $-$1.177 (0.442) & 0.043 & 0.326 \\
    Including 2022 & $-$1.910 & $-$2.200 & $-$1.178 & 0.058 & 0.320 \\
    \bottomrule
  \end{tabular}
\end{table}

The fitted parameters of the GG distribution int he
GAMLSS framework in Equations~\eqref{eq:gg}--\eqref{eq:sigma} are
summarized in Table~\ref{tab:ggfit}. Results obtained from both
excluding and including 2022 data are reported. The fixed-effect
parameters include $\beta_0$, $\gamma_0$, and $\nu$, corresponding to
the intercept of the log-location, log-scale, and shape of the GG
distribution, respectively. Random effects account for variability at
the venue level $\sigma_v$ on the log-scale of the $\mu$ parameter
and at the heat level $\sigma_h$ on the log-scale of the $\sigma$
parameter in the density in Equation~\eqref{eq:gg}. The variance
of the venue-level random effect is smaller than the heat-level random
effect variance, suggesting that heat-level variability in the scale
parameter is substantial, though on the dispersion parameter. When the
2022 data is included, the fixed-effect estimates remain stable,
indicating robustness, but the venue-level random-effect variance
increases to 0.0576, reflecting additional variability in the location
parameter, while the heat-level random-effect variance remains stable.
decreases slightly to 0.3198. These results
highlight that RTs are influenced by both venue and heat-level
factors, with heat-level variability playing a more prominent role in the scale
parameter, and that the inclusion of 2022 introduces greater venue-level
variability, likely due to systematic differences in RTs that year.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{diagnosis.pdf}
  \caption{Diagnosis of the fitted GG distribution with
    random effects in model parameters: kernerl density of 1 million
    observations drawn from the fitted model overlaid with the
    histogram of the observed RTs (left); Q-Q plot of the
    normal z-score of the quantile residuals from the fitted model.}
  \label{fig:diagnosis}
\end{figure}

Figure~\ref{fig:diagnosis} presents diagnostic checks for the fitted
GG distribution model with random effects. The left
panel compares the kernel density estimate of one million simulated reaction
times from the fitted model to the histogram of the observed RTs.
The close alignment between the density curve and the histogram suggests that
the fitted GG model adequately captures the overall distribution of the
RTs. The right panel shows a Q-Q plot of the z-scores of the
quantile residuals from the fitted model. The points lie approximately along
the 45-degree reference line, indicating that the residuals are consistent
with the standard normal distribution, supporting the adequacy of the model
fit. These diagnostics collectively demonstrate that the fitted model provides
a reasonable representation of the observed RT data.

@@

The fitted GG GAMLSS model with both venue- and heat-level
random effects provides a framework for assessing how extreme RTs
below certain thresholds are. The probability of observing a RT
below a given threshold, assuming no intentional false starts, was approximated
by generating 10 million realizations from the fitted model.
Table~\ref{tab:Sim_probability} summarizes the probabilities of observing
RTs below 0.08, 0.09, and 0.10 seconds under two scenarios: one
excluding and the other including data from 2022. Excluding 2022 slightly
reduces the probability of observing a fast RT, but the difference
is small. For example, the probability of a RT below 0.10 seconds
decreases from $2.76 \cdot 10^{-3}$ (approximately one in 362 starts) to
$1.94 \cdot 10^{-3}$ (approximately one in 515 starts) when 2022 is excluded.
Lowering the RT threshold from 0.10 to 0.08 seconds drastically
reduces the likelihood of observing a RT below the barrier, with
the probability dropping from one in every 362 starts (at 0.10 seconds) to one
in every 14620 starts (at 0.09 seconds) and one in every 146198 starts (at 0.08
seconds) when 2022 is included. These results highlight the rarity of extremely
fast RTs and substantiate the recommendations of \citet{komi2009iaaf}
to carefully consider the selection of RT thresholds.
